Inferring true changes in microbial abundance from taxonomically-biased microbiome measurements

### 2022-01-14 abundance-measurement.Rmd
Revise treatment of main theoretical results
This commit revises the language, notation, and explanation of the main
theoretical results about the effect of bias. Changes include
- using 'abundance' instead of 'density' throughout
- updating equations and text to use the new notation and language
  develeped in the appendix, in which the set S of all species used in
  the denominator of a proportion calculation is explicitly defined, and
  the mean efficiency is written as the efficiency of the multi-species
  set S.
- clarifying what is meant by 'absolute abudance'
- revising the abundance-me

To understand how bias affects the measured differential abundance between samples, we first consider how it affects measurements of individual samples.

## Model of MGS measurement

Our primary tool for understanding the impact of taxonomic bias on MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the *compositional* nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
We consider a set of microbial species that are measured by a specific MGS protocol that extracts, sequences, and bioinformatically analyzes samples to assign reads to various microbial species.
We assume that assigned reads have been correctly assigned to the given species and sample.
Thus we ignore the possibility of contamination or false-positive taxonomic assignment, though we allow that many reads may not receive a taxonomic assignment and so ultimately be discarded.
Unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the 'noise' or random error due to the random sampling of sequencing reads and other aspects of the MGS process.

Our model stipulates that the assigned read count of species $i$ in sample $a$ equals its cell density multiplied by a species-specific factor and a sample-specific factor,
\begin{align}
  (\#eq:measurement-model)
  \text{reads}_{i}(a)
  = \text{density}_{i}(a) \quad \cdot
    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
    \cdot \quad
    \underbrace{\text{sequencing effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
\end{align}
The species-specific factor is the *relative measurement efficiency* (or *efficiency* for short) of the species, which represents how much more easily that species is measured (converted from cells to assigned reads) relative to an abitrarily chosen reference species (@mclaren2019cons).
The sample-specific factor, which we call the *sequencing effort* of the sample, reflects the fact that the number of reads per unit cell density varies among samples even in the absence of taxonomic bias due to variation in total density, library normalization, and total sequencing-run output.
Equation \@ref(eq:measurement-model) implies that the total reads from all species in the sample equal
\begin{align}
  (\#eq:total-reads)
  \text{total reads}(a)
    = \text{total density}(a) \cdot \text{mean efficiency}(a) \cdot \text{sequencing effort}(a),
\end{align}
where 
\begin{align}
  (\#eq:mean-efficiency)
  \text{mean efficiency}(a) 
  \equiv \frac{\sum_{j}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{total density}(a)}
\end{align}
is the average or mean efficiency of cells in the sample.

## Relative abundances (proportions and ratios)

We distinguish between two types of species-level *relative abundances*.
The first type consists of the *ratios* among the abundances of two or more species, such as a ratio of 5:4:2 among three species.
The second type consists of the *proportion* of an individual species, equal to its abundance divided by that of a particular root taxon (e.g. Prokaryotes).
DA analyses of proportions consider how the proportion of a species changes indepently of other species, whereas DA analyses of ratios consider the conserted change in relative abundance among two or more species.
Although proportion-based analysis is dominant, microbiome researchers increasingly use ratio-based analyses adapted from the statistical framework of Compositional Data Analysis (CoDA).
The different manner in which taxonomic bias affects ratios vs. proportions (@mclaren2019cons) forms the basis in subsequent sections for understanding the extent to which its effects cancel in DA analysis.

The standard measurement of the proportion of species $i$ in sample $a$ is given by the ratio of its read count to the total,
\begin{align}
  (\#eq:prop-meas)
  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{total reads(a)}}.
\end{align}
From Equations \@ref(eq:measurement-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas), it follows that
\begin{align}
  (\#eq:prop-error)
  \widehat{\text{prop}}_{i}(a)
  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)},
\end{align}
where $\text{prop}_{i}(a) = \text{density}_{i}(a) / \text{total density}(a)$ is the actual proportion.
Equation \@ref(eq:prop-error) says that taxonomic bias creates a multiplicative error equal to the ratio of the species' efficiency to the mean efficiency in the sample.
Consequently, the same species can be over- or under-measured depending on the composition of the given sample (Figure \@ref(fig:error-proportions)).
For instance, in samples from two hypothetical communities in Figure \@ref(fig:error-proportions), Species 3 has an efficiency of 6 and is under-measured in Sample 1 (which has a mean efficiency of 8.33) but over-measured in Sample 2 (which has a mean efficiency of 3.15).

The measured ratio between species $i$ and $j$ is given by the ratio of their read counts, 
\begin{align}
  (\#eq:ratio-meas)
  \widehat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)}.
\end{align}
From Equations \@ref(eq:measurement-model) and \@ref(eq:ratio-meas), it follows that 
\begin{align}
  (\#eq:ratio-error)
  \widehat{\text{ratio}}_{i/j}(a)
%  \frac{\text{reads}_{i}(a)}{\text{reads}_{j}(a)}
  &= \frac{\text{density}_{i}(a)}{\text{density}_{j}(a)} \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}.
\end{align}
The fold error in the ratio between two species simply equals the ratio in their efficiencies and is therefore constant across samples.
For instance, in the example of Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.

@@

(ref:cap-error-proportions) **Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.** Top row: Error in proportions measured by MGS in two microbiome samples that contain different relative abundances of three species. Bottom row: Error in the measured fold-change in the third species that is derived from these measurements.

@@

**Difference between species proportions and species ratios:**
Consistent taxonomic bias at the species level creates a consistent fold error in the ratios among species, but a varying fold error in the proportions of individual species.
But a proportion is, after all, just a ratio between two taxa in which the denominator taxon is the entire taxonomic domain under study.
The difference in behavior from that of ratios between species arises because the efficiency of higher-level taxa, which consist of the aggregate of many species, varies across samples with the changing relative abundances among its constituent species.

## Absolute abundances (cell densities)

A wide range of experimental methods exist for converting the relative-abundance information in MGS measurements into measurements of (absolute) cell densities;
however, these methods can be generally partitioned into two types, based on whether the absolute-abundance information derives from knowing the density of the total community or knowing the density of one or more individual species.

**Methods using the density of the total community:**
The total microbial density of a community can be assayed by a variety of methods (reviewed in Appendix \@ref(review-absolute-methods)).
Once the total density is known, the density of individual species can be estimated by multiplying the total density by the species' proportions (as measured by MGS),
\begin{align}
  (\#eq:density-prop-meas)
  \widehat{\text{density}}_{i}(a) 
  &= \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{total density}}(a)
\\&= \text{reads}_{i}(a) \cdot \frac{\widehat{\text{total density}}(a)}{\text{total reads}(a)}.
\end{align}
We can rewrite the species' density measurements to account for the effect of taxonomic bias on the MGS proportions (Equation \@ref(eq:prop-error)) and any error in the total density measurement, 
\begin{align}
  (\#eq:density-prop-error)
  \widehat{\text{density}}_{i}(a) 
  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)} 
%  \cdot \frac{\widehat{\text{total density}}(a)}{\text{total density}(a)}.
  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{total density}}(a) \end{array}.
\end{align}
The variable fold error that bias creates in the measured proportions directly transfers to the density measurements.

Taxonomic bias can also create systematic error in the total-density measurements (Appendix \@ref(appendix-total-density)).
For now, we suppose that the total density can be measured accurately; 
later, Section \@ref(solutions) describes how experiments might be designed such that taxonomic bias in the total density measurement offsets that in the MGS proportions, leading to a (more) consistent fold error in the measured densities.

**Methods using the density of one or more reference species:**
The density of one or more species can be directly measured using targeted measurement methods like species-specific qPCR; alternatively, the densities of one or more species may be thought to be constant across samples either from background knowledge or because they have been spiked in at a fixed density (these situations are reviewed in Appendix \@ref(review-absolute-methods)).
These _reference species_ with known or constant density can be used to convert read counts to densities for all species.
For instance, given the density of a reference species $r$, we can estimate the density of a focal species $i$ by 
\begin{align}
  (\#eq:density-ratio-meas)
%  \widehat{\text{density}}_{i}(a) = \frac{\text{reads}_{i}(a)}{\text{reads}_{r}(a)} \cdot \widehat{\text{density}}_{r}(a).
  \widehat{\text{density}}_{i}(a) = \text{reads}_{i}(a) \cdot \frac{\widehat{\text{density}}_{r}(a)}{\text{reads}_{r}(a)}.
\end{align}
Alternatively, if we only know that the reference species is constant, we can set $\widehat{\text{density}}_{r}(a)$ in Equation \@ref(eq:density-ratio-meas) equal to 1 to estimate the density in unknown but fixed units, which is sufficient for multiplicative DA analysis.
(Appendix \@ref(model-details) considers possible extensions to multiple reference species.)
It follows from Equation \@ref(eq:ratio-error) that the error caused by taxonomic bias in this density measurement is
\begin{align}
  (\#eq:density-ratio-error)
  \widehat{\text{density}}_{i}(a) 
  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}} 
  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{density}_r}(a) \end{array}.
\end{align}
Here, it is the constant error in the measured ratio of species $i$ to $r$ that propagates into the density measurement.
There will generally be systematic error in the density of the reference species; however, if the systematic fold error is constant across samples, so will be that of the densities of other species.

**Differences between the two approaches:**
Equations \@ref(eq:density-prop-error) and \@ref(eq:density-ratio-error) show that the manner in which taxonomic bias in the sequencing measurement impacts the two approaches mirrors that of proportions and ratios.
If the fold error in the total-density or reference species measurement is negligible (or at least constant) then the fold errors in species densities by the total-density approach vary across samples with the mean efficiency, whereas the fold errors by the reference-species approach are constant. 
The fundamental determinant of whether constant error is obtained is not which of these two equations is ultimately used, but rather whether densities of individual species, which (we assume) have constant efficiency, or the total community, which has varying efficiency, provide the absolute-abundance information.
Studies using spike-ins as constant reference species will often, instead of using Equation \@ref(eq:density-ratio-meas), first use the spike-in species to estimate the total density, and then estimate the density of focal species with Equation \@ref(eq:density-prop-meas).
If the calculation is done carefully, however, this method still yields constant fold errors so long as the species efficiencies (including those of the spike-in) are constant across samples (Appendix \@ref(appendix-total-density)).

>>>>

We begin by considering how taxonomic bias affects the relative- and absolute-abundance measurements which serve as the input to microbiome DA analyses.

Our primary tool for understanding the impact of taxonomic bias on MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
This model is the simplest that respects the multiplicative nature of taxonomic bias and the *compositional* nature of MGS measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
We consider a set of microbiome samples measured by a specific MGS protocol that extracts, sequences, and bioinformatically analyzes to taxonomically assign reads to a set of microbial species $S$.
Various forms of taxonomic assignment are possible; for simplicity, we suppose that reads are assigned to the species level, with reads that cannot be uniquely assigned being discarded.
We ignore the possibility that reads are incorrectly assigned to the wrong species or sample.
Unless otherwise stated, we treat the sequencing measurement as deterministic, ignoring the 'random' variation in read counts that arise from the sampling of sequencing reads and other aspects of the MGS process.

In our model, the assigned read count of a species $i$ in a sample $a$ equals its abundance multiplied by species-specific and sample-specific factors,
\begin{align}
  (\#eq:measurement-model)
  \text{reads}_{i}(a)
  = \text{abun}_{i}(a) \quad \cdot
    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
    \cdot \quad
    \underbrace{\text{effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
\end{align}
The species-specific factor, $\text{efficiency}_{i}$, equals the *relative measurement efficiency* (or simply *efficiency*) of the species---how much more easily that species is measured (converted from cells to assigned reads) relative to a arbitrary fixed reference species (@mclaren2019cons).
We assume the efficiencies of particular species are consistent across samples.
The variation in efficiency among species corresponds to the taxonomic bias of the MGS protocol.
The sample-specific factor, $\text{effort}(a)$, we call the *sequencing effort* for that sample; it captures the variation in the total number of assigned reads due to experimental features such as library normalization and total sequencing-run output.
Equation \@ref(eq:measurement-model) implies that the total number of assigned reads in sample $a$ equals
\begin{align}
  (\#eq:total-reads)
  \text{reads}_S(a)
    = \text{abun}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{effort}(a),
\end{align}
where $\text{abun}_{S}(a) \equiv \sum_{j\in S}\text{abun}_j(a)$ is the total abundance of species in $S$ and 
\begin{align}
  (\#eq:mean-efficiency)
  \text{efficiency}_S(a) 
    \equiv \frac{\sum_{j\in S}\text{abun}_j(a) \cdot \text{efficiency}_j}{\text{abun}_S(a)}
\end{align}
is the _mean efficiency_ over all species in $S$.
<!-- The subscript $S$ serves as a reminder that many reads may not be assigned---for example, reads that fail to pass quality control or that come from the host or other taxonomic domains than what is being studied. -->

## Relative abundance (proportions and ratios)

We distinguish between two types of species-level *relative abundances* within a sample.
The *proportion* of species $i$ in sample $a$ equals to its abundance relative to the total abundance of all species in $S$,
\begin{align}
  (\#eq:prop)
  \text{prop}_{i}(a) 
    &\equiv \frac{\text{abun}_i(a)}{\text{abun}_S(a)}.
%  \\&= \frac{\text{abun}_i(a)}{\sum_{i \in S}\text{abun}_i(a)}.
\end{align}
The *ratio* between two species $i$ and $j$ equals the abundance of $i$ relative to that of $j$,
\begin{align}
  (\#eq:ratio)
  \text{ratio}_{i/j}(a) = \frac{\text{abun}_i(a)}{\text{abun}_j(a)}.
\end{align}
Proportions and ratios each form the basis for popular DA methods;
ratio-based methods are commonly referred to as Compositional Data Analysis (CoDA) methods.

Taxonomic bias has a different effect on species proportions versus species ratios.
The proportion of a species is typically measured by its proportion of assigned reads,
\begin{align}
  (\#eq:prop-meas)
  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{reads}_S(a)}.
\end{align}
We can rewrite the right-hand side (using Equations \@ref(eq:measurement-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas)) to find
\begin{align}
  (\#eq:prop-error)
  \widehat{\text{prop}}_{i}(a)
  &= \text{prop}_{i}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_S(a)}}_{\substack{\text{variable} \\ \text{fold error}}}.
\end{align}
Taxonomic bias creates a multiplicative error in the species' proportion equal to its efficiency relative to the mean efficiency in the sample.
Consequently, the species's proportion is measured as too high in samples that are dominated by species with lower efficiencies, and measured as too low in samples that are dominated by species with higher efficiencies.
This phenomenon is illustrated in two hypothetical communities in Figure \@ref(fig:error-proportions). 
Species 3 has an efficiency of 6; it is under-measured in Sample 1, which has a mean efficiency of 8.33, but over-measured in Sample 2, which has a mean efficiency of 3.15.

The measured ratio between species $i$ and $j$ is given by the ratio of their read counts, 
\begin{align}
  (\#eq:ratio-meas)
  \widehat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)}.
\end{align}
From Equations \@ref(eq:measurement-model) and \@ref(eq:ratio-meas), it follows that 
\begin{align}
  (\#eq:ratio-error)
  \widehat{\text{ratio}}_{i/j}(a)
%  &= \frac{\text{abun}_{i}(a)}{\text{abun}_{j}(a)} 
  &= \text{ratio}_{i/j}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}}_{\substack{\text{constant} \\ \text{fold error}}}.
\end{align}
Taxonomic bias creates a multiplicative measurement error in the ratio that is equal to the ratio in their efficiencies; the error is therefore constant across samples.
For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.

@@

(ref:cap-error-proportions) **Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.** Top row: Error in proportions measured by MGS in two hypothetical microbiome samples that contain different relative abundances of three species. Bottom row: Error in the measured fold-change in the third species that is derived from these measurements. Species' proportions may be measured as too high or too low depending on sample composition. For instance, Species 3 has an efficiency of 6 and is under-measured in Sample 1 (which has a mean efficiency of 8.33) but over-measured in Sample 2 (which has a mean efficiency of 3.15).

@@

The fact that the multiplicative error in proportions varies while that in ratios is constant forms the basis for understanding the effect of bias on different methods for measuring absolute abundances from MGS and the extent to which bias cancels in different DA analyses.

## Absolute abundance

Researchers often would like to know how the _absolute abundance_ of a species changes.
In this context, _absolute_ means not relative to other species; typically, the abundance of interest is relative to non-microbial entities such the volume, mass, or amount of host DNA in the sample.
A wide range of experimental methods have been used to convert the relative abundances from MGS measurements into measurements of absolute abundances.
Yet there has so far been little consideration as to how the resulting measurements are affected by taxonomic bias in the MGS measurements or in the supplemental measurements, such as flow cytometry and qPCR, that they often involve.
Here and in Appendix \@ref(review-absolute-methods) we describe how various methods for measuring absolute abundance are predicted to be affected by taxonomic bias, with an eye towards determining whether the measurement error is variable or constant.

Our discussion supposes the 'abundance' of interest is cell number, though it applies to other quantities, such as biomass and genome copy number, that may be more relevant in different biological contexts.

We consider two general classes of methods: Those in which absolute-abundance information is derived from measurement of the aggregate abundance of the total community, and those in which it drives from targeted measurement (or prior knowledge) of one or more particular species.

**Normalization by total abundance:**
Measurements of species absolute abundances are often obtained by making a (non-MGS) measurement of total community abundance, equating it with the aggregate abundance of the species $S$ measured by MGS, and multiplying this total by the proportions from MGS,
\begin{align}
  (\#eq:density-prop-meas)
  \widehat{\text{abun}}_{i}(a) 
  &= \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{abun}}_S(a).
\end{align}
This measurement is affected by taxonomic bias in the MGS measurement as well as in the total-abundance measurement.
For example, measurement of total community abundance by 16S qPCR is affected by variation among species in extraction efficiency, 16S copy number, and PCR binding and amplification bias.
If the species-level efficiencies of the total-abundance measurement are constant across samples and we neglect other sources, then we can express the total-abundance measurement as
\begin{align}
  (\#eq:total-density-error)
  \widehat{\text{abun}}_S(a) 
  &= \sum_{i\in S} \text{abun}_i(a) \cdot \text{efficiency}^{\text{tot}}_i
\\&= \text{abun}_S(a) \cdot \text{efficiency}^{\text{tot}}_S(a),
\end{align}
where $\text{efficiency}_{i}^{\text{tot}}(a)$ is the absolute measurement efficiency of species $i$ for the total-abundance measurement and $\text{efficiency}^{\text{tot}}_S(a)$ is the mean efficiency of the total-abundance measurement in the sample.
<!-- Note, we have assumed that the total-abundance and MGS measurements are sensitive to the same species -->
The equations \@ref(eq:total-density-error) and \@ref(eq:prop-error) for the error in total-abundance and proportion measurements imply that the species abundance measurement \@ref(eq:density-prop-meas) has error given by
\begin{align}
  (\#eq:density-prop-error)
  \widehat{\text{abun}}_{i}(a) 
  = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i} \cdot \text{efficiency}^{\text{tot}}_S(a)}{\text{efficiency}_S(a)}.
\end{align}

Equation \@ref(eq:density-prop-error) indicates that the multiplicative error in the measured absolute abundance of a species equals its MGS efficiency relative to the mean MGS efficiency in the sample, multiplied by the mean efficiency of the total measurement.
As in the case of proportions (Equation \@ref()), the error depends on sample composition through the two mean efficiency terms and so may vary across samples.
On the other hand, if the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, the two can offset and lead to a relatively stable error.
We discuss how this possibility might apply to real experimental workflows below.

**Normalization by one or more reference species:**
Suppose we had a measurement of the (absolute) abundance of a _reference species_ $r$.
In the absence of taxonomic bias, all species are expected to have the same ratio of reads to abundance in a sample (Equation \@ref(eq:measurement-model)).
Thus the abundance-to-reads ratio for species $r$ can serve as a conversion factor allowing us to obtain the abundance of an arbitrary species $i$,
\begin{align}
  (\#eq:density-ratio-meas)
  \widehat{\text{abun}}_{i}(a) = \text{reads}_{i}(a) \cdot \frac{\widehat{\text{abun}}_{r}(a)}{\text{reads}_{r}(a)}.
\end{align}
The abundance of one or more reference species can be directly measured using targeted measurement methods like species-specific qPCR and used with Equation 
\@ref(eq:density-ratio-meas) to obtain absolute abundance for all species.
To our knowledge, this approach has not previously been suggested.
<!-- (Appendix \@ref(model-details) considers possible extensions to multiple reference species.) -->
Instead, reference-based measurement has been used in the context of spike-in experiments or normalization to a species (such as the host) which is treated as having a constant abundance.
In a spike-in experiment, the abundance of reference species are added in known (up to experimental error) abundances to each sample.
When normalizing to a species that is treated as having a constant but unknown abundance, we set $\widehat{\text{abun}}_{r}(a)$ in Equation \@ref(eq:density-ratio-meas) equal to 1 and interpret the abundances as having fixed but unknown units, which is sufficient for multiplicative DA analysis.

The error in the abundance measurement \@ref(eq:density-ratio-meas) due to taxonomic bias in the MGS measurement is given by
\begin{align}
  (\#eq:density-ratio-error)
  \widehat{\text{abun}}_{i}(a) 
  = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}}
  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{abun}_r}(a) \end{array}.
\end{align}
The constant error in the measured ratio of species $i$ to $r$ (see Equation \@ref(eq:ratio-error)) propagates to the abundance measurement.
There will generally also be systematic error in the abundance of the reference species; however, if the systematic fold error is constant across samples, so will be that of the abundances of other species.

Spike-ins are instead sometimes used to measure absolute abundances using Equation \@ref(eq:density-prop-meas).
In this case, an intermediate step is taken in which the total community abundance is first measured by the ratio of non-spike-in to spike-in reads.
<!-- CHECK -->
If done correctly, this calculation yields results that are identical to directly applying Equation \@ref(eq:density-ratio-meas) (Appendix \@ref(total-density-ref)).

**Difference between the two approaches:**
Reference-species normalization yields constant fold errors because it is based on species-level read counts and abundances and we assume that efficiencies are constant at the species level.
In contrast, total-abundance normalization is based on aggregates of species (for the calculation of proportion and the total-abundance measurement) and so depends on mean efficiencies, which can vary across samples.
